---
title: "Relationship between miles per gallon (MPG) and transmission"
author: "Jonathan Lok-Chuen Lo"
date: "November 21, 2015"
output: pdf_document
geometry: margin=1in
---

```{r set_options, echo=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, 
    warning = FALSE, 
    message = FALSE,
    fig.path = "figure/",
    fig.height = 3,
    fig.width = 6,
    fig.pos = "center")
```

# Executive Summary

# Explorary Data Analysis

Load the *mtcars* dataset and the required libraries. Transmission is converted
into a factor variable. Boxplot of MPG vs transmission is shown in Appendix A. 

```{r load_data}
library(ggplot2); library(dplyr); library(datasets)
data(mtcars)
mtcars$am = factor(mtcars$am, c(0, 1), labels = c("Automatic", "Manual"))
```

In the first glance, manual transmission seems to be more efficient in terms of
MPG. However, MPG is highly dependent on other factors such as engine power and
weight. Appendix B and C shows MPG vs number of cylinders and weight
respectively, between automatic and manual transmission. It shows the dataset
contains a larger number of smaller manual transmission cars.

# Model Selection

As an initial trial, let us start with fitting the MPG vs the transmission type
only. 

```{r}
fit <- lm(mpg ~ am, data = mtcars)
summary(fit)$coef
```

Results above show that the average MPG for automatic transmission is 
`r round(fit$coef[1], 2)`, while manual transmission is `r round(fit$coef[2], 2)`
higher. P-values show the results are significant using 5% confidence level.

However, as discussed earlier, MPG is dependent on many other factors. Two
important factors are the car weight and the engine power. For engine power, 1/4
mile time, displacement and horsepower are all good indicators. To avoid 
including unnecessary variables, a nested model testing is performed.

```{r model_selection}
fit1 <- lm(mpg ~ am, data = mtcars) 
fit2 <- update(fit, mpg ~ am + qsec + wt) 
fit3 <- update(fit, mpg ~ am + qsec + wt + hp + disp) 
anova(fit1, fit2, fit3)
```

From the ANOVA result above, using 1/4 mile time and weight as regressors 
reduce the residual sum of squares (RSS) and it is statistically significant.
Including horsepower and displacement only improves the RSS marginally, and also
the improvement is not significant. Hence, the regression model is chosen to use
transmission (as factor variable), 1/4 mile time and weight. 

# Regression Model

First, a regression model is fitted by assuming regressors 1/4 mile time and 
weight are having the same slopes for both manual and automatic transmissions.

```{r model_fit}
fitSameSlope <- lm(mpg ~ am, data = mtcars)
summary(fitSameSlope)$coef
```

```{r model_fit1}
fitSameSlope <- lm(mpg ~ am + wt + qsec, data = mtcars)
summary(fitSameSlope)$coef
```

With this assumption, results above shows that the average MPG for manual 
transmission does seem to have an improved MPG, or 
`r round(fitSameSlope$coef[2], 3)` MPG higher as compared to automatic 
transmissions of average `r round(fitSameSlope$coef[1], 3)` MPG.

However, as discussed earlier, the dataset contains more data for manual 
transmission cars having smaller weight with less engine power, as compared to
automatic transmission. To further investigate, a regression model using
different slopes for the regressors is used.

```{r model_fit2}
fitDifferentSlope <- lm(mpg ~ am * (wt + qsec), data = mtcars)
summary(fitDifferentSlope)$coef
```

Above shows that although the average MPG for manual transmission is still
higher, it is no longer statically significant. In addition, the slope of weight
against MPG is significant larger for manual transmission, or slope is 
`r round(fitDifferentSlope$coef[5], 3)` MPG per 0.001lb from that of automatic 
transmission. This suggests that weight of cars might have different impacts on 
MPG as weight increases. 

A residual plot is shown in Appendix D. The fitted values against residual does
not contain any obvious pattern. However, the Q-Q plot seems to suggest the
data might not be normal, which seems to coincide our observations that 
dependency of MPG against weight is skewed.

# Conclusion

In conclusion, 

## Appendix A - Plot of MPG vs Transmission

```{r echo = FALSE}
g <- ggplot(mtcars, aes(am, mpg, fill = am))
g <- g + geom_boxplot()
g <- g + labs(x = "Transmission", y = "Miles per Gallon (MPG)")
g
```

## Appendix B - Plot of MPG vs Number of Cylinders for Different Transmission

```{r echo = FALSE}
g <- ggplot(mtcars, aes(factor(cyl), mpg, fill = am))
g <- g + geom_boxplot()
g <- g + labs(x = "Number of Cylinder", y = "Miles per Gallon (MPG)")
g <- g + scale_fill_discrete(name = "Transmission")
g
```

## Appendix C - Plot of MPG vs Weight for Different Transmission

```{r echo = FALSE}
g <- ggplot(mtcars, aes(wt, mpg))
g <- g + geom_point(aes(color = am))
g <- g + labs(x = "Weight (lb/1000)", y = "Miles per Gallon (MPG)")
g <- g + scale_fill_discrete(name = "Transmission")
g
```

## Appendix D - Residual Plot of the Proposed Regression Model

```{r echo = FALSE, fig.height = 6}
par(mfrow = c(2, 2))
plot(fitDifferentSlope)
```


