---
title: "Regression Model Course Project : Motor Trend"
author: "Jonathan Lok-Chuen Lo"
date: "November 21, 2015"
output: pdf_document
geometry: margin=1in
---

```{r set_options, echo=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, 
    warning = FALSE, 
    message = FALSE,
    fig.path = "figure/",
    fig.height = 4,
    fig.width = 6,
    fig.pos = "center")
```

## Executive Summary

This report aims to explore the relationship between a set of variables and 
miles per gallon (MPG). In particular, we are interested in: 1) Is an automatic 
or manual transmission better for MPG: 2) Quantify the MPG difference between 
automatic and manual transmissions. Data used in the analysis was based on the 
*mtcars* dataset which was extracted from the 1974 Motor Trend US magazine for
32 automobiles (1973-74 models).

Regression of MPG against only transmission types shows that manual transmission
is 7.245 MPG better than automatic transmission. However, by including weight
and 1/4 mile time, the result becomes inconclusive. It is found that the dataset
contains more manual transmission cars of smaller size and lower engine power,
which could skew the analysis (i.e. the apparent better MPG of manual
transimission could be due to other factors such as weight and engine power).

## Explorary Data Analysis

Load the *mtcars* dataset and the required libraries. Transmission is converted
into a factor variable. Box-plot of MPG vs transmission is shown in Appendix A. 

```{r load_data}
library(ggplot2); library(dplyr); library(datasets)
data(mtcars)
mtcars$am = factor(mtcars$am, c(0, 1), labels = c("Automatic", "Manual"))
```

In the first glance, manual transmission seems to be more efficient in terms of
MPG. However, MPG is highly dependent on other factors such as engine power and
weight. Appendix B and C shows MPG vs number of cylinders and weight
respectively, between automatic and manual transmission. It shows the dataset
contains a larger number of smaller manual transmission cars.

## Model Selection

MPG is dependent on many other factors. Two important factors are the car weight
and engine power. For engine power, 1/4 mile time, displacement and
horsepower are all good indicators. To avoid including unnecessary variables, a
nested model testing is performed.

```{r model_selection}
fit1 <- lm(mpg ~ am, data = mtcars) 
fit2 <- update(fit1, mpg ~ am * (qsec + wt)) 
fit3 <- update(fit1, mpg ~ am * (qsec + wt + hp + disp))
anova(fit1, fit2, fit3)
```

From the ANOVA above, using 1/4 mile time and weight as regressors reduce the 
residual sum of squares (RSS) and the result is significant. Including 
horsepower and displacement only improves the RSS marginally. Hence, the 
regression model is chosen to use transmission , 1/4 mile time and weight.

## Regression Model

First, we run a regression model between MPG and transmission types only.

```{r model_fit}
fitSameSlope <- lm(mpg ~ am, data = mtcars)
summary(fitSameSlope)$coef
```

Results above shows that the average MPG for manual transmission is 
`r round(fitSameSlope$coef[2], 3)` MPG higher as
compared to automatic transmissions of average 
`r round(fitSameSlope$coef[1], 3)` MPG.

However, as discussed earlier, the dataset contains more data for manual 
transmission cars having smaller weight with less engine power, as compared to
automatic transmission. A regression with adjustment using weight and 1/4 mile
time is used.

```{r model_fit2}
fitDifferentSlope <- lm(mpg ~ am * (wt + qsec), data = mtcars)
summary(fitDifferentSlope)$coef
```

Above shows that although the average MPG for manual transmission is still
higher, it is no longer statically significant. In addition, the slope of weight
against MPG is significant larger for manual transmission, or slope is 
`r round(fitDifferentSlope$coef[5], 3)` MPG per 0.001lb from that of automatic 
transmission. This suggests that weight of cars might have different impacts on 
MPG as weight increases. 

A residual plot is shown in Appendix D. The fitted values against residual does
not contain any obvious pattern. However, the Q-Q plot seems to suggest the
data might not be normal, which seems to coincide our observations that 
dependency of MPG against weight is skewed.

## Conclusion

In summary, the dataset does not provide an conclusive answer to whether 
automatic or manual transmission is better for MPG. This is because the dataset 
contains more manual transmission cars of small size and lower engine power. To
properly answer the question, additional randomized data is required.

## Appendix A - Plot of MPG vs Transmission

```{r echo = FALSE}
g <- ggplot(mtcars, aes(am, mpg, fill = am))
g <- g + geom_boxplot()
g <- g + labs(x = "Transmission", y = "Miles per Gallon (MPG)")
g
```

## Appendix B - Plot of MPG vs Number of Cylinders for Different Transmission

```{r echo = FALSE}
g <- ggplot(mtcars, aes(factor(cyl), mpg, fill = am))
g <- g + geom_boxplot()
g <- g + labs(x = "Number of Cylinder", y = "Miles per Gallon (MPG)")
g <- g + scale_fill_discrete(name = "Transmission")
g
```

## Appendix C - Plot of MPG vs Weight for Different Transmission

```{r echo = FALSE}
g <- ggplot(mtcars, aes(wt, mpg))
g <- g + geom_point(aes(color = am))
g <- g + labs(x = "Weight (lb/1000)", y = "Miles per Gallon (MPG)")
g <- g + scale_fill_discrete(name = "Transmission")
g
```

## Appendix D - Residual Plot of the Proposed Regression Model

```{r echo = FALSE, fig.height = 6}
par(mfrow = c(2, 2))
plot(fitDifferentSlope)
```


